---
title: "22.10.22_PU5058_REPORT.Rmd"
output: pdf_document
date: "2022-10-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, tidy.opts=list(width.cutoff=80), tidy=TRUE) 
```


## Aim

The aim is to highlight to local government, schools and parents the reduction in mental well being scores for girls in S4, specifically in rural areas. 

***

## Load Packages
```{r install packages}

#install of the following packages, if they are not already installed, this is done in the console as you only need to do it once. 
#The # in front of the code would need removed first, as I have left it as a comment as I have already installed all of the packages.
#Information about each of the packages is commented in the load libraries section below.

#install.packages("tidyverse") #remember the inverted commas around the name of the package

#install.packages("here")

#install.packages("sf")

#install.packages("mapview")

#install.packages("cowplot")

#install.packages("leafpop")
```

***

## Libraries 
```{r load the libraries within the packages, message=FALSE, warning=FALSE}
#load various packages needed to create the visualisations


#here package is used to create file paths from a top level directory(folder)
library(here)

#tidyverse package has lots of functions useful for Data Science
library(tidyverse)

#sf (simple features) package is used to work with shapefiles, that includes spatial vector data, that are required 
#for the creation of maps
library(sf)

#flexdashboard package is used to display multiple visualisations. I only have one map to display so this is not required. 
#In the written discussion I mention that a table of additional information to accompany the map would be beneficial and then 
#this package would be required.
#library(flexdashboard)

#mapview package has multiple functions that are used to create interactive visualisations of spatial data
library(mapview)

#gt package has functions to create tables, such as add headers, change column width. This could be used as an 
#improvement as discussed in the written report.
#library(gt)

#cowplot is used in addition to the graphing package ggplot (part of the tidyverse package) that offers functions to 
#create publication quality visualisations. Examples are choose themes, annotate graphs and align plots.
library(cowplot)

#plotly is used to create interactive graphs to be displayed on the internet
library(plotly)

#this is not a package to be installed but this sets the theme of the map, including the overall font size and the font 
#size of labels and the title, line width.
theme_set(theme_cowplot())

#Provides functions to edit the pop up table from the interactive map.
library(leafpop)




```

There are multiple sources of information about different packages, one source can be found [here (1).](https://rdocumentation.org/)
Default settings for cowplot can be found [here (2).](https://rdrr.io/cran/cowplot/man/theme_cowplot.html)

***

## Read in the data

```{r read in the data, message=FALSE, warning=FALSE}

#Information about where the data came from can be found in the accompanying report.

#to read in the data we will use the readr package which was loaded within the tidyverse package
#We will assign the data files to objects by giving them names

#These files include multiple variables including the mean wellbeing score (measure) and the year and 
#gender of the group (indicator)
S2_Boys<-read_csv(here("Input/22.10.22_01_PU5058_REPORT_S2_BOYS.csv"))
S2_Girls<-read_csv(here("Input/22.10.22_02_PU5058_REPORT_S2_GIRLS.csv"))
S4_Girls<-read_csv(here("Input/22.10.22_03_PU5058_REPORT_S4_GIRLS.csv"))
S4_Boys<-read_csv(here("Input/22.10.22_04_PU5058_REPORT_S4_BOYS.csv"))

#These files are required for creating the map. They include data zone information and shape files.

#read in the data for the council areas providing all multiple zone areas for each council in Scotland
simd_indicators <- read_csv(here("Input/SIMD2020v2_indicators.csv"))

#read in the geospatial data (shape files) used to create the map. Note: quiet=T results in no messages 
#being displayed during upload
datazone_sf <- st_read(here("Input/sc_dz_11.shp"), quiet = T)

```
***

## Prepare the data

### Bar Chart

The first preparation of the data is for the bar chart.

```{r Investigate S2_Boys, S2_Girls, S4_Boys, S4_Girls}
#check the variables are the same in each file
head(S2_Boys)
head(S2_Girls)
head(S4_Boys)
head(S4_Girls)
```

```{r join the files}
#join the files for S2, S4 boys and girls into one dataset
combined_data<-bind_rows(S2_Boys,S4_Boys,S2_Girls,S4_Girls,.id = NULL)

#it can be seen in the environment panel that the combined_data dataset has 100 observations, which is 
#what is expected as the 4 individual datasets had 25 observations each.


```
The variables that we are interested in are indicator (giving year and gender of the pupils), area_name (council area) and measure(mean wellbeing score). Further information about how the mean wellbeing score is calculated can be found in the accompanying report.

We need to know which council areas are in the S4_Girls dataset. This information will be useful when preparing the data for the map.


```{r check councils in S4_Girls}
S4_Girls$area_name%>%
  table()
```
There are 24 councils with mean wellbeing score data.

The combined data has a variable called indicator which includes the gender and the school year. This is not tidy data so the indicator column is split into School Year and Gender. The variable Gender_School_Year which includes both variables is also created but to only be used as a label on the bar chart x-axis.

```{r create new variables - tidy data}



#create a new dataset to store the new variables
combined_data_substr<-combined_data%>%
#filter for only Scotland, this will result in 4 observations to be plotted on the bar chart
  filter(area_type=="Scotland")%>%
#Create a new variable and select only the 1st to 7th character
  mutate(Gender_School_Year=substr(indicator,1,7))%>%
#Create a new variable and select only the 1st to 2nd character
  mutate(School_Year=substr(indicator,1,2))%>%
#Create a new variable and select only the 4th to 7th character
  mutate(Gender=substr(indicator,4,7))
  
#check the new columns have been created
head(combined_data_substr)
 
  
```

The data is now only 4 observations and it can be seen that there are no missing values or unusual entries. No further investigation is required. The required variables will be selected when the bar chart is created.

***

### Map

The preparation for the map can now be completed.

The map will require more than 4 observations for the mean wellbeing score. It will require one for each of the council areas. We will use the original S4_Girls data set for the map.

```{r select variables and rename a column S4_Girls}
#Only the variables measure (mean Well being Score) and area_name (council area) are needed for the map
#Create a new dataset to save the changes
S4_Girls_col_rename<-S4_Girls%>%
  #select the variables needed to create the graph
  select(measure,area_name)%>%
  #rename the column measure to be more meaningful when plotted in the map
  rename(Average_Wellbeing_Score=measure)
```


The data zones to create the map will be sourced from a dataset that includes information from the Scottish Index of Multiple Deprivation report. Further information can be found in the accompanying report.

```{r investigate simd_indicators}
#Investigate the variables in the simd indicators dataset
head(simd_indicators)
```
The variables of interest will be Data_Zone and Council_area. Scotland is split into 6,976 geographic data zones.

```{r check for missing values}
#check for missing values in the simd_indicators data set
simd_indicators%>%
summarise_all(~sum(is.na(.)))
```


```{r check for ununusal character %}
#check for unusual characters in the simd_indicators dataset(*Acknowledgments)
simd_indicators%>%
  select(Data_Zone, Council_area)%>%
  filter_all(any_vars(str_detect(.,pattern ="%")))
```

```{r check for unusual characters \\*}
#check for unusual characters in the simd_indicators dataset*(Acknowledgements)
simd_indicators%>%
  select(Data_Zone, Council_area)%>%
  filter_all(any_vars(str_detect(.,pattern ="\\*")))
```
```{r check which councils are in the simd_indictors}
#Check which council areas are listed within the simd_indicators dataset
simd_indicators$Council_area%>%table()

```
This information was checked because some of the council areas were not included in the research for the mean wellbeing scores.The number of times each council appears represents how many data zones make up the council area.

```{r create a new dataset, select columns, filter for data for map}
#create a new dataset to include only the variables Data_Zone and Council-area.
simd_selected_col<-simd_indicators%>%
  #select the variables needed to be able to join to the S4_Girls_col_rename dataset
  select(Data_Zone, Council_area)%>%
  #remove the council areas that have no mean wellbeing score
  filter(Council_area !="Scottish Borders" & Council_area !="West Dunbartonshire" & Council_area !="Shetland Islands" & Council_area !="West Lothian" & Council_area !="South Ayrshire" & Council_area !="South Lanarkshire" & Council_area !="Renfrewshire" & Council_area !="Stirling")
```

There are no missing values, unusual characters or council areas with no mean wellbeing score in the data set.

The council areas in both data sets simd_selected_col (area_name) and in the S4_Girls_col_rename (Council_area) can now be used to join the datasets. This will create one dataset with the mean wellbeing scores and data zone information. This dataset will then be combined to the datazone shape files needed to create the map.




```{r join data sets council area}
#updated version use this joining councils 
S4_councils_DZ<-left_join(simd_selected_col,S4_Girls_col_rename, by=c("Council_area"="area_name"))
```

```{r investigate S4_councils_DZ}
head(S4_councils_DZ)
```

The datazone shapefile dataset will now be checked.

```{r investigate datazone_sf}
#Investigate the variables in the datazone_sf dataset
head(datazone_sf)
```

```{r check for missing values datazone_sf}
#datazone_sf%>%
  #summarise_all(~sum(is.na(.)))
# I tried to check for missing values in the datazone_sf but I received a message "no loop for break/next, 
#jumping to top level. This would need further investigation. I assume it is to do with being a shapefile.
```

The shape files now need to be joined using the data zones.
```{r join data using data zones}
#Add shape files by joining the data by the data zones.

S4_councils_DZ_sf<-left_join(datazone_sf, S4_councils_DZ, by=c("DataZone"="Data_Zone"))
```

```{r investigate S4_councils_DZ_sf}
#Investigate the variables in the S4_councils_DZ_sf dataset
head(S4_councils_DZ_sf)
```

The dataset S4_councils_DZ_sf now contains the shape files for each datazone, council area and the mean wellbeing score that will be used to create the map.

***

## Create Visualisation

### Column Chart

The column chart can now be created.

```{r create bar chart, message=FALSE, warning=FALSE}
#Please note the format of the comments and code below were used to run an RMarkdown file without the interactive map.
#Select the dataset that we want to use to create the graph
  #Select the variables (columns) that we will use to create the graph
  #Reorder the columns so Gender_School_Year variable is put in order by descending measure 
  #variable on the x-axis. The y-axis is the variable measure and we are going to choose the colour of the column depending on the gender.
  #Define which colours you want the columns to be, boys will be deep sky blue and girls pink.
  #A column chart is chosen instead of a bar because we already have the total measure values.
  #Remove the legend, we have the gender on the x-axis for people who are colour blind. Centre the title of the graph
  #Add an x-axis label
  #Add a y-axis label
  #Add a title
combined_data_substr %>% select(Gender_School_Year, Gender, measure)%>% ggplot(aes(x=reorder(Gender_School_Year,-measure),y=measure, fill=Gender))+geom_col()+scale_fill_manual(values=c("deepskyblue","pink"))+theme(legend.position = "none", plot.title =element_text(hjust=0.5))+xlab("High School Year and Gender")+ylab("Average Wellbeing Score")+ggtitle("Average Wellbeing Scores for Pupils in High School")
    
```



The colours were chosen from a large selection available in R, [colours in R (3).](https://r-graph-gallery.com/42-colors-names.html)


***

### Map

The map can now be created.

```{r plot map}
#Define the colour palette to use when creating the map. The argument direction = -1 has not been used 
#because the low numbers are to have dark colours rather than vice versa. The number of colours required is 7. 
pal = viridisLite::rocket(n = 7)

 #Create a new object to store the selected variables for the map
S4_councils_DZ_sf %>%

#Select the relevant variables to create the map
select(DataZone,Council_area,Average_Wellbeing_Score) %>%

#mapview is a package that can quickly create interactive maps
  mapview(

#Select the type of maps to use
    map.types = "OpenStreetMap",

  #Define which variable will be used to determine the colour shades of the polygons
   zcol = "Average_Wellbeing_Score", 
  
#The council areas will be visible when the mouse hovers over an area on the map
   label = S4_councils_DZ_sf$Council_area,

#Select the variables that you want to be visible in the pop up table in the interactive map
   popup=popupTable(S4_councils_DZ_sf, zcol=c("Council_area", "Average_Wellbeing_Score")),
    
 #Define the name of the layer that we want to show on the map
    layer.name = "Average_Wellbeing_Score",

 #Set the opacity of the colour fills to 0.8, 1 is fully opaque (you can't see through). The boundaries of the council areas have to be visible. 
   alpha.regions = 0.8,

 #This defines that pal (defined above) will be the colour pallette used to fill the polygons in the map. Break points have been defined so only 7 colours are required.
   col.regions = pal,at=c(45,45.5,46,46.5,47,47.5,48)
    )
```

Further information about the [virisisLite (4).](https://cran.r-project.org/web/packages/viridisLite/viridisLite.pdf) package can found using the link.


## References

https://rdocumentation.org/ (1)

https://rdrr.io/cran/cowplot/man/theme_cowplot.html (2)

https://r-graph-gallery.com/42-colors-names.html (3)

https://cran.r-project.org/web/packages/viridisLite/viridisLite.pdf (4)

## Acknowledgements
https://jessbutler.github.io/simd/ *(You can see the code and data under "Sources".)


